Tidy data structure to support exploration and modeling of temporal data

### 2018-12-07 main.Rmd
restarted introduction

Temporal-context data, $X_{jt}$, consist of $j = 1, \dots, N_{i}$ observational units indexed at different time points, $1 \le t \le T$. Time forms the contextual basis of temporal data, but it can arrive in many possible formats, which makes preparing it for modeling and forecasting painful. For example, data can be recorded at various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information in addition to time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures. The data problems are grouped into two types of analysis, time series and longitudinal.

```{r pipeline-ann}
include_graphics("img/time-series-structure.pdf")
```

@@

## Time series and longitudinal data

Despite of exactly the same data input, the representation of time series and longitudinal data diverges due to different modelling approaches.

Time series can be univariate or multivariate, and for modelling requires relatively large in length (i.e. large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterising temporal dynamics. The time series supporting modeling are represented as vectors or matrices in most statistical software, with the standards being provided by the R [@R-base] packages `ts`, **zoo** [@zoo2005], and **xts** [@R-xts] (*Why include MATLAB [@Matlab]??*). Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wide format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric variables. In addition, time indexes are stripped off the data and implicitly inferred as attributes or meta-information. It strictly requires that the number of observations must be the same across all the series. Data wrangling from the form that data arrives in, to this specialist format can be frustrating and difficult, inhibiting the variety of downstream tasks such as analytics.

@@

Longitudinal data representation has been implemented in Stata's time series module [@Stata] and R package **plm** [@plm2008]. The underpinning data structure is a two-dimensional column-homogeneous array, as other tabular data. Specifying a longitudinal data set from a generic array needs explicitly declaring time and individuals (or panel variable in Stata's `tsset` command). Therefore, each row of the data can be identified by a specific individual and time point. Individuals, however, can be only declared through a single variable, not multiple.

@@

@viewing-pipeline describes a viewing pipeline for interactive statistical graphics, that takes control of the transformation from data to plot. @xgobi, @ggobi, @orca, @plumbing and @xie2014reactive implemented data pipelines for the interactive statistical software **xgobi**, **ggobi**, **orca**, **plumbr** and **cranvas**, respectively. The pipeline is typically described with a one way flow, from data to plot. For interactive graphics, where all plots need to be updated when a user interacts with one plot, the events typically trigger the data pipeline to be run. @xie2014reactive uses a reactive programming framework, to implement the pipeline, in which user's interactions trigger a sequence of modules to update their views, that is, practically the same as running the data pipeline producing each plot.

@@

A table to list the verbs and what they do.

* **implicit missingness**: `has_gaps()`, `count_gaps()`, `fill_gaps()`
* **row-wise**: `filter()`, `slice()`, `arrange()`
* **column-wise**: `mutate()`, `select()`, `summarize()`
* **group-wise**: `index_by()`, `group_by()`
* **reshape**: `gather()`, `spread()`, `nest()`, `unnest()`

@@

As a special case of data frame, a tsibble pipes into the grammar of graphics straight way, making most use of this powerful graphical system. It should be easy to create and extend some specialist plotting methods based on tsibble structure, such as autocorrelation plots [@tsibblestats] and calendar-based graphics [@sugrrants].

@@

data-centric & human-centered.

@@

Time representation could be arbitrary, for example R's native `POSIXct` and `Date` for versatile date-times, nano time for nanosecond resolution [@R-nanotime], and pure numbers in simulations. Yet ordered factors can also be a source of time, such as month names from January to December and weekdays from Monday to Sunday. Tsibble supports an extensive range of index types from numerics to nano time, but there might be custom indices used for some occasions, for example school semesters. These academic terms vary from one institute to another within an academic year, which is defined differently from a calendar year. New index would be immediately recognized by the software upon defining `index_valid()`, as long as it can be ordered from past to future. The interval regarding semesters is further outlined through `pull_interval()`. As a result, the rest software methods such as `has_gaps()`, `count_gaps()` and `fill_gaps()` will have instant support for data that contains this new index.

@@

```{r map-prep}
```

<!-- 
```{r map-airlines, fig.cap = "(ref:map-airlines-cap)"}
```

(ref:map-airlines-cap) An overview of US airline traffic in 2017. Points indicate locations of airports, and lines the connecting routes across the country. There are a few hubs connecting most airlines and sporadic airports.
-->

The dataset of 2017 on-time performance for US domestic flights represents event-driven data caught in the wild, sourced from US Bureau of Transportation Statistics (reference). It contains `r format(NROW(flights), big.mark = ",")` operating flights with many measurements (such as departure delay, arrival delay in minutes, and other performance metrics) and detailed flight information (such as origin, destination, plane number and etc.) in a tabular format. This kind of event describes each flight scheduled for departure at a time point in its local time zone. Every single flight would be uniquely identified by the flight number and its scheduled departure time, from a passenger's point of view. In fact, it fails to pass the tsibble hurdle due to duplicates in the original data. An error is immediately raised when attempting to convert this data into a tsibble, and a closer inspection has to be carried out to locate the issue. The **tsibble** package provides tools to easily locate the duplicates in the data with `duplicates()`. Below shows the problematic entries.

@@

<!--

```{r dl771, fig.cap = "(ref:dl771-cap)"}
```

(ref:dl771-cap) Flight DL771 operates on multiple routes between different origins and destinations. It suggests that the key required for identification consists of flight number and origin.

-->

@@

<https://data.gov.au/dataset/smart-grid-smart-city-customer-trial-data>

@@

```{r write-bib}
write_bib(c("dplyr", "tidyr"), file = "rpkgs.bib")
```

>>>>

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $1 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. An overview of data science workflow is given in Figure \ref{fig:pipeline-ann}. The divide between data that feeds into the beginning of analysis and the current time series standard that fits into modelling is significant, making data preparation for modeling and forecasting painful. For example, data can be recorded at various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors.

```{r pipeline-ann, fig.cap = "(ref:pipeline-ann-cap)"}
include_graphics("img/time-series-structure.pdf")
```

(ref:pipeline-ann-cap) An overview of data science program. The existing standards are model-focused, resulting misalliance between data and model.

@@

## Time series and longitudinal data {#sec:time-series}

The data problems are grouped into two types of analysis, time series and longitudinal. Despite of exactly the same data input, the representation of time series and longitudinal data diverges due to different modelling approaches.

Time series can be univariate or multivariate, and for modelling requires relatively large in length (i.e. large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. The time series supporting modeling are represented as vectors or matrices in most statistical software. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wide format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric variables. In addition, time indexes are stripped off the data and implicitly inferred as attributes or meta-information. It strictly requires that the number of observations must be the same across all the series. Data wrangling from the form that data arrives in, to this specialist format can be frustrating and difficult, inhibiting the variety of downstream tasks such as analytics.

@@

Longitudinal data representation has been implemented. The underpinning data structure is a two-dimensional column-homogeneous array, as other tabular data. Specifying a longitudinal data set from a generic array needs explicitly declaring time and individuals (or panel variable in Stata's `tsset` command). Therefore, each row of the data can be identified by a specific individual and time point. Individuals, however, can be only declared through a single variable, not multiple.

@@

@viewing-pipeline describes a viewing pipeline for interactive statistical graphics, that takes control of the transformation from data to plot. @xgobi, @ggobi, @orca, @plumbing and @xie2014reactive implemented data pipelines for the interactive statistical software **Xgobi**, **Ggobi**, **Orca**, **plumbr** and **cranvas**, respectively. The pipeline is typically described with a one way flow, from data to plot. For interactive graphics, where all plots need to be updated when a user interacts with one plot, the events typically trigger the data pipeline to be run. @xie2014reactive uses a reactive programming framework, to implement the pipeline, in which user's interactions trigger a sequence of modules to update their views, that is, practically the same as running the data pipeline producing each plot.

@@

\begin{table}[!hbt]
\caption{A test}\label{tab:verb-table}
\centering
\begin{tabular}{ lll }
\toprule
 & Verb & Description \\ \midrule
\multirow{3}{*}{Time gaps} & \texttt{has\_gaps()} & Test if a tsibble has gaps in time \\
 & \texttt{count\_gaps()} & Count and report time gaps \\
 & \texttt{fill\_gaps()} & Fill in gaps by values and functions \\ \hline
\multirow{4}{*}{Row-wise} & \textbf{\texttt{filter()}} & Pick rows based on conditions \\
 & \texttt{filter\_index()} & Provide a shorthand for time subsetting \\
 & \textbf{\texttt{slice()}} & Select rows based on row positions \\
 & \textbf{\texttt{arrange()}} & Sort the ordering of row by variables \\ \hline
\multirow{4}{*}{Column-wise} & \textbf{\texttt{select()}} & Pick columns by variables \\
 & \textbf{\texttt{mutate()}} & Add new variables \\
 & \textbf{\texttt{transmute()}} & Add new variables \\
 & \textbf{\texttt{summarise()}} & Aggregate values over time \\ \hline
\multirow{3}{*}{Group-wise} & \texttt{index\_by()} & Group by new time index \\
 & \textbf{\texttt{group\_by()}} & Group by one or more variables \\
 & \texttt{group\_by\_key()} & Group by key variables \\ \hline
\multirow{4}{*}{Reshape data} & \textbf{\texttt{gather()}} & David Batty \\
 & \textbf{\texttt{spread()}} & Eirik Bakke \\
 & \textbf{\texttt{nest()}} & Eirik Bakke \\
 & \textbf{\texttt{unnest()}} & Jody Morris \\
\bottomrule
\end{tabular}
\end{table}

@@

As a special case of data frame, a tsibble pipes into the grammar of graphics straight way, making most use of this powerful graphical system. It should be easy to create and extend some specialist plotting methods based on tsibble structure, such as autocorrelation plots and calendar-based graphics [@wang-calendarviz-2018].

@@

Time series are represented as matrices, with standards being provided by native `ts` object in R, extended by **zoo** [@zoo2005], and **xts** [@R-xts]. As discussed in Section \ref{sec:time-series}, this data organization sets far apart from data origin. Due to time isolated from the main data, the supporting functions that intend to deal with time must be applied to whole data. That data-centric & human-centered.

@@

Time representation could be arbitrary, for example R's native `POSIXct` and `Date` for versatile date-times, nano time for nanosecond resolution implemented in **nantotime** [@R-nanotime], and pure numbers in simulations. Yet ordered factors can also be a source of time, such as month names from January to December and weekdays from Monday to Sunday. Tsibble supports an extensive range of index types from numerics to nano time, but there might be custom indices used for some occasions, for example school semesters. These academic terms vary from one institute to another within an academic year, which is defined differently from a calendar year. New index would be immediately recognized by the software upon defining `index_valid()`, as long as it can be ordered from past to future. The interval regarding semesters is further outlined through `pull_interval()`. As a result, the rest software methods such as `has_gaps()`, `count_gaps()` and `fill_gaps()` will have instant support for data that contains this new index.

@@

The dataset of 2017 on-time performance for US domestic flights represents event-driven data caught in the wild, sourced from US Bureau of Transportation Statistics [@flights-data]. It contains `r format(NROW(flights), big.mark = ",")` operating flights with many measurements (such as departure delay, arrival delay in minutes, and other performance metrics) and detailed flight information (such as origin, destination, plane number and etc.) in a tabular format. This kind of event describes each flight scheduled for departure at a time point in its local time zone. Every single flight would be uniquely identified by the flight number and its scheduled departure time, from a passenger's point of view. In fact, it fails to pass the tsibble hurdle due to duplicates in the original data. An error is immediately raised when attempting to convert this data into a tsibble, and a closer inspection has to be carried out to locate the issue. The **tsibble** package provides tools to easily locate the duplicates in the data with `duplicates()`. Below shows the problematic entries.

@@

@@

@@

The **tsibble** package provides an elegant solution to manage and manipulate medium-sized temporal data in memory.

```{r write-bib}
write_bib(c("dplyr", "tidyr", "nanotime"), file = "rpkgs.bib")
```
